### 2023-03-16 paper.tex
final touches

\begin{minipage}{0.7\textwidth}
\small
\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution.
Analyzing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
Typically, this is simply done by dividing the genome into large tiles and averaging in these.
We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions to quantify methylation in, and a better quantitation method than simple averaging.
We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to detect biologically meaningful regions near 
Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
\end{minipage}
\vspace{6.5ex}
}]

@@

The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.
of counts of distinct RNA molecules), one row for each cell and one column for each gene.
A first goal is usually to assign cell types or states to cells.
To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm \todo{is this even correct?}.
Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts.
Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components.
As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on a large number of genes.
Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score.
Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

\begin{figure*}
    \begin{center}
    \includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
    \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
    Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(C)} \todo{plot this as discussed, as in Fig.~6D} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
    Data from \citet{kremer_scnmt}.}
    \label{fig:correlation}
\end{figure*}

@@

\subsection{Mean neighbor score} \label{methods:score}
To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$.
The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.

\subsection{Correlation of DNA methylation and gene expression}
To assess correlation between gene expression and the methylation status of VMRs or promoters, we
first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-treshold 0.05}.
We then quantified DNA methylation at VMRs and promoters with \citettt{scbs matrix}.
We defined promoters as $\pm2000$~bp regions centered on a genes transcription start site (TSS).
When multiple TSSs were annotated, we chose the TSS of the "principal" isoform \citep{appris}.
Log-normalized expression values reported by \citet{kremer_scnmt} were then correlated with methylation of the genes promoter, or with methylation of the VMR closest to the gene body.
When multiple VMRs intersected the gene body, we chose the VMR with the highest methylation variance.
As a measure of methylation we used the shrunken mean of the residuals.
We omitted lowly expressed genes (with scRNA-seq counts in $\lt5\%$ of cells) and promoters and VMRs with low scBS coverage (in $\lt10$ cells).

>>>>

\begin{minipage}{0.7\textwidth}
\small
\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution.
Analyzing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
Typically, this is simply done by dividing the genome into large tiles and averaging in these.
We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions to quantify methylation in, and a better quantitation method than simple averaging.
We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to detect biologically meaningful regions near genes involved in core functions of specific cell types.
Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
\end{minipage}
\vspace{6.5ex}
}]

@@

The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.
of counts of distinct RNA molecules), one row for each cell and one column for each gene.
A first goal is usually to assign cell types or states to cells.
To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) \todo{is this even correct?} to the normalized fractions before taking the logarithm.
Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts.
Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components.
As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on a large number of genes.
Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score.
Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

\begin{figure*}
    \begin{center}
    \includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
    \textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
    Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
    Data from \citet{kremer_scnmt}.}
    \label{fig:correlation}
\end{figure*}

@@

\subsection{Mean neighbor score} \label{methods:score}
To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$, rounded to the nearest integer.
The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.

\subsection{Correlation of DNA methylation and gene expression}
To assess correlation between gene expression and the methylation status of VMRs or promoters, we
first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-treshold 0.05}.
We then quantified DNA methylation at VMRs and promoters with \texttt{scbs matrix}.
We defined promoters as $\pm2000$~bp regions centered on a genes transcription start site (TSS).
When multiple TSSs were annotated, we chose the TSS of the "principal" isoform \citep{appris}.
Log-normalized expression values reported by \citet{kremer_scnmt} were then correlated with methylation of the gene's promoter, or with methylation of the VMR closest to the gene body.
When multiple VMRs intersected the gene body, we chose the VMR with the highest methylation variance.
As a measure of methylation we used the shrunken mean of the residuals.
We omitted lowly expressed genes (with scRNA-seq counts in $<5\%$ of cells) and promoters and VMRs with low scBS coverage (in $<10$ cells).
